Towards a Practical Face Recognition System: \\ Robust Alignment and Illumination by Sparse Representation

### 2011-03-29 faces_system.tex
Changes from Zihan

CMU Multi-PIE provides the most extensive test set among public datasets. This
database contains images of 337 subjects across simultaneous variation in pose,
expression, and illumination.  Of these 337 subjects, we use all of the 249
subjects present in Session 1 as the training set. The remaining 88 subjects
are treated as ``impostors'', or invalid images. For each of the 249 training
subjects, we include frontal images of 7 frontal illuminations,\footnote{They
are illuminations $\{0,1,7,13,14,16,18\}$ of \cite{Gross2008-FGR}. For each
directional illumination, we subtract the ambient-illuminated image 0.} taken
with neutral expression. As suggested by the work of
\cite{Georghiades2001-PAMI}, these extreme frontal illuminations would be
sufficient to linearly represent other frontal illuminations, as will also be
corroborated by the next experiment on our own dataset. For the test set, we
use all 20 illuminations from Sessions 2-4, which were recorded over a period
of several months. The dataset is challenging due to the large number of
subjects, and due to natural variation in subject appearance over time.  Line 5
of Table \ref{tab:MultiPIE-recognition} shows the result of our recognition
pipeline on each of the 3 testing sessions. Our algorithm achieves recognition
rates above $90\%$ for all three sessions.  For the test images, our iterative
alignment was initialized automatically via the Viola and Jones' face detector.

\begin{table}
\caption{Recognition rates on the Multi-PIE database for different pairings of alignment
and recognition stages.}
\vspace{-.2in}
\centerline{
\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c| }
\hline
\backslashbox{Rec.}{Align.} 
& \multicolumn{3}{|c|}{Face Detector} 
& \multicolumn{3}{|c|}{Manual} 
& \multicolumn{3}{|c|}{Iterative} 
\\
\hline
Session $\rightarrow$	& 2		&3			&4			& 2		&3			&4			& 2		&3			&4		\\
\hline
NS	& 30.8\%	& 29.4\%	& 24.6\%	& 77.6\%	& 74.3\%	& 73.4\%	& 84.5\%	& 82.3\%	& 81.4\% \\
\hline
NN	& 26.4\%	& 24.7\%	& 21.9\%	& 67.3\%	& 66.2\%	& 62.8\%	& 73.5\%	& 69.6\%	& 69.3\% \\
\hline
LDA	& 5.1\%		& 5.9\%		& 4.3\%		& 49.4\%	& 44.3\%	& 47.9\%	& 91.0\%	& 89.9\%	& 88.1\% \\
\hline
LBP	& 39.9\%	& 38.1\%	& 33.9\%	& 93.3\%	& 91.2\%	& 92.9\%	& {\bf 95.2\%}	& {\bf 94.7\%}	& {\bf 93.5\%} \\
\hline
SRC	&			&			&			&			&			&			& 93.9\%	& 93.8\%	& 92.3\% \\
\hline
\end{tabular}
\label{tab:MultiPIE-recognition} }
\end{table}

@@

Table \ref{tab:MultiPIE-recognition} shows that the
recognition rates on Multi-PIE indeed increase with this new
window. In addition, Figure \ref{fig:failed-examples}(a), (b),
and (c) illustrate three representative subjects for which the
recognition rates of our algorithm are significantly boosted
with the new window. However, we should mention that the best choice of the window
is problem-specific and there is
not a simple guideline to follow. For example, although the new
window performs better on Multi-PIE, the same window does not
help at all on our own database, which will be introduced in
the next section. This is because most of the training and
testing images in our database are taken on the same day so the
variation in hair style is very small. Hence, excluding the
forehead part may actually result in loss of useful
discriminative information.

\subsection{Comparison to Existing Work}
% Classical algorithms
We compare our result to baseline linear-projection-based
algorithms, such as Nearest Neighbor (NN), Nearest Subspace
(NS) \cite{Lee2005-PAMI}, and Linear Discriminant Analysis
(LDA) \cite{Belhumeur1997-PAMI}.\footnote{We do not list
results on PCA \cite{Turk1991-CVPR} as its performance is
always below that of Nearest Subspace.} Since these algorithms
assume pixel-accurate alignment, they are not expected to work
well if the test image is not well aligned with the training. In
Table~\ref{tab:MultiPIE-recognition}, we report the results of
these classical algorithms with three types of testing image alignment: 
1.\ alignment from the Viola and Jones' detector,
2.\ alignment via manually selected outer eye corners, and 
3.\ the output of our iterative alignment algorithm.
The performance drop of the LDA algorithm on Multi-PIE reported here
seems to agree with that reported already in
\cite{Gross2008-FGR}.  All of the classical algorithms benefit
greatly from being paired with our iterative alignment algorithm.

% LBP
Finally, we compare our result to Local Binary Patterns (LBP)
\cite{Ahonen2006-PAMI}, a local appearance descriptor which is able to capture
fine details of facial appearance and texture.  Due to its robustness to
variations in illumination, facial expression, aging and other changes, LBP has
achieved the state-of-the-art face recognition performance in the scenario when
only one sample per person is used for training \cite{Tan06facerecognition}. In
this paper, we follow the same steps as in \cite{Ahonen2006-PAMI} to construct
an LBP descriptor for each training and testing sample. The $80\times 60$ face
region is first divided into a regular $10\times 10$ grid of cells, each of
size $8\times 6$ pixels. Within each cell, the histogram of 59 uniform binary
patterns is then computed, where the patterns are generated by thresholding 8
neighboring pixels in a circle of radius 2 using the central pixel value.
Finally, the local histograms are concatenated to produce the global descriptor
vector. As suggested in \cite{Ahonen2006-PAMI}, the recognition is performed
using a nearest neighbor classifier with Chi square distance as the distance
measure and we report the recognition rates with the same two types of input as
before.  Recognition results for LBP on Multi-PIE are shown in
Table~\ref{tab:MultiPIE-recognition}.  We attribute the improvement in
performance of LBP over SRC in this experiment to its robustness to
illumination components that are not in the training set.  Some of the test
images used in this experiment are taken under illuminations not present in the
training subset.  Note that, like the other classical algorithms, the
performance of LBP is still sensitive to misalignment. This is notable given
that LBP is often applied without any special alignment.

\begin{table}
\caption{Recognition rates on the Multi-PIE database for \cite{Yang2010-CVPR} and variants
of Algorithm 1}
\vspace{-.2in}
\centerline{
\begin{tabular}{|c|c|c|c|c| }
\hline
Recognition rate & Session 2 & Session 3 & Session 4  \\
\hline
\cite{Yang2010-CVPR} & {\bf 95.2}\% & 93.4\% & 95.1\% \\
\hline
{Algorithm 1, $S=1$} & 90.7\% & 89.6\% & 87.5\% \\
\hline
{Algorithm 1} & 93.9\% & 93.8\% & 92.3\% \\
\hline
{Algorithm 1 with improved window} & 95.0\% & {\bf 96.3}\% & {\bf 97.3}\% \\
\hline
\end{tabular}
\label{tab:MultiPIE-recognition2} }
\vspace{-.3in}
\end{table}

To demonstrate that the sparse representation based recognition step is indeed
beneficial even when there are no impostors, we include results for recognition
based only on the alignment error residuals(i.e. $S=1$), shown in 
Table \ref{tab:MultiPIE-recognition2}, row 6.

We also compare our result to the recent work
\cite{Yang2010-CVPR}. Notice that in \cite{Yang2010-CVPR}, the
initial registration is again obtained from manually selected
outer eye corners. Then, a supervised hierarchical sparse
coding model based on local image descriptors is trained, which
enjoys certain translation invariant properties. With the same
training and testing sets, \cite{Yang2010-CVPR} is able to
handle any remaining misalignment and achieves state-of-the-art
performance on the CMU Multi-PIE database.
Table~\ref{tab:MultiPIE-recognition2} shows that our algorithm
achieves similar or better performance on different sessions of
Multi-PIE.

Although our algorithm is not designed for
recognition when there is only a single gallery image per user,
we compare its
performance with LBP within this setting for completeness.
For this experiment, we use the FERET dataset
\cite{phillips1998feret}, which contains five standard
partitions: `fa' is the gallery containing 1196 frontal images
of 1196 subjects, and `fb', `fc', `dup1' and `dup2' are four
sets of probe images. The testing sets differ from the training
in facial expression (`fb'), illumination (`fc'), aging (`dup1'
) and long aging (`dup2'). In fact, except for `fb', we notice
significant changes of illumination in all the other three test
sets. For the training, we again crop and normalize the face
region from each original image to an $80\times 60$ window
using manually marked eye coordinates \cite{Deng2010-PR}. In
Table~\ref{tab:FERET-recognition}, we report the performance of
our algorithm on the four test sets, with input directly
obtained from the Viola and Jones' detector, and the
performance of LBP with input obtained from both face detector
and manual alignment using the eye coordinates.

@@

Figure \ref{fig:roc-multipie} plots the receiver operating characteristic (ROC)
curves, which are generated by sweeping the threshold $t$ through the entire
range of possible values for each algorithm.\footnote{Rejecting invalid images
not in the entire database is much more difficult than deciding if two face
images are the same subject. Figure 8 should not be confused with typical ROC
curves for face similarity, e.g., \cite{PhillipsP2007}.} On the left we can see
that the SCI based recognition approach significantly outperforms the other
algorithms, including LBP, even when all algorithms are coupled with our
proposed iterative alignment.  In the right plot we again see that classical
algorithms, and even LBP, are very sensitive to alignment.  Similar contrasts
between our algorithm and baseline algorithms were also observed for SRC in
\cite{Wright2009-PAMI}, though on much smaller datasets.
\begin{figure}
\centerline{
	\begin{tabular}{cc}
	\includegraphics[width=3in]{figures_pami/pami_roc_revision2.png} &
	\includegraphics[width=3in]{figures_pami/pami_roc2.png} \\
	{\small With iterative alignment} & {\small Classical Algorithms with manual alignment} 
\\
	\end{tabular}
}
\caption{{\bf ROC curves} for subject validation on Multi-PIE database, (a) with iterative alignment,
and (b) for the classical algorithms with manual alignment}\label{fig:roc-multipie}
%\vspace{-.6in}
\end{figure}

@@

\begin{figure}[t]
\centering
\includegraphics[scale=0.75,clip=true]{figures_pami/multipie_occ/occ10.png}
\includegraphics[scale=0.75,clip=true]{figures_pami/multipie_occ/occ20.png}
\includegraphics[scale=0.75,clip=true]{figures_pami/multipie_occ/occ30.png}
\includegraphics[scale=0.75,clip=true]{figures_pami/multipie_occ/occ40.png}
\includegraphics[scale=0.75,clip=true]{figures_pami/multipie_occ/occ50.png}
\vspace{-.25in}
\caption{{\bf Recognition under varying level of
random block occlusion.} The above row shows examples of occluded test images with occlusion level from 10\% to 50\%. Our method maintains high recognition rates up to 30\% occlusion:} 
\vspace{3mm}
{\small
\begin{tabular}{|c|c|c|c|c|c| }
\hline
Percent occluded & 10\% & 20\% & 30\% & 40\% & 50\%  \\
\hline
\hline
Recognition rate & 99.6\% & 94.9\% & 79.6\% & 46.5\% & 19.8\% \\
\hline
\end{tabular}
}
\label{fig:multipie-occ-rec}
\vspace{-.25in}
\end{figure}

>>>>

CMU Multi-PIE provides the most extensive test set among public
datasets. This database contains images of 337 subjects across
simultaneous variation in pose, expression, and illumination.
Of these 337 subjects, we use all of the 249 subjects present
in Session 1 as the training set. The remaining 88 subjects are
treated as ``impostors'', or invalid images. For each of the
249 training subjects, we include frontal images of 7 frontal
illuminations,\footnote{They are illuminations
$\{0,1,7,13,14,16,18\}$ of \cite{Gross2008-FGR}. For each
directional illumination, we subtract the ambient-illuminated
image 0.} taken with neutral expression. As suggested by the
work of \cite{Georghiades2001-PAMI}, we choose these extreme
frontal illuminations in the hope that they would linearly
represent other frontal illuminations well. For the test set,
we use all 20 illuminations from Sessions 2-4, which were
recorded over a period of several months. The dataset is
challenging due to the large number of subjects, and due to
natural variation in subject appearance over time. Table
\ref{tab:MultiPIE-recognition2} shows the result of our
algorithm on each of the 3 testing sessions. Our algorithm
achieves recognition rates above $90\%$ for all three sessions.
For the test images, our iterative alignment was initialized
automatically via the Viola and Jones' face detector. To
demonstrate that the sparse representation based recognition
step is indeed beneficial even when there are no impostors, we
include results for recognition based only on the alignment
error residuals (i.e. $S=1$), shown in row 1.

\begin{table}
\caption{Recognition rates on the Multi-PIE database for
Algorithm 1 and \cite{Yang2010-CVPR}} \vspace{-.2in}
\centerline{
\begin{tabular}{|c|c|c|c|c| }
\hline
Recognition rate & Session 2 & Session 3 & Session 4  \\
\hline
{Algorithm 1, $S=1$} & 90.7\% & 89.6\% & 87.5\% \\
\hline
{Algorithm 1} & 93.9\% & 93.8\% & 92.3\% \\
\hline
{Algorithm 1 with improved window} & 95.0\% & {\bf 96.3}\% & {\bf 97.3}\% \\
\hline
\cite{Yang2010-CVPR} & {\bf 95.2}\% & 93.4\% & 95.1\% \\
\hline
\end{tabular}
\label{tab:MultiPIE-recognition2} }
\end{table}

@@

Table \ref{tab:MultiPIE-recognition2} shows that the
recognition rates on Multi-PIE indeed increase with this new
window. In addition, Figure \ref{fig:failed-examples}(a), (b),
and (c) illustrate three representative subjects for which the
recognition rates of our algorithm are significantly boosted
with the new window. However, we should mention that the best
choice of the window is problem-specific and there is not a
simple guideline to follow. For example, although the new
window performs better on Multi-PIE, the same window does not
help at all on our own database, which will be introduced in
the next section. This is because most of the training and
testing images in our database are taken on the same day so the
variation in hair style is very small. Hence, excluding the
forehead part may actually result in loss of useful
discriminative information.

\subsection{Comparison to Existing Work}
We first compare our result to the recent work
\cite{Yang2010-CVPR}. Notice that in \cite{Yang2010-CVPR}, the
initial registration is obtained from manually selected outer
eye corners. Then, a supervised hierarchical sparse coding
model based on local image descriptors is trained, which enjoys
certain translation invariant properties. With the same
training and testing sets, \cite{Yang2010-CVPR} is able to
handle the remaining misalignment and achieves state-of-the-art
performance on the CMU Multi-PIE database.
Table~\ref{tab:MultiPIE-recognition2} shows that our algorithm
achieves similar or better performance on different sessions of
Multi-PIE.

% Classical algorithms
To better examine the effectiveness of our iterative alignment
algorithm, we next compare our result to baseline
linear-projection-based algorithms, such as Nearest Neighbor
(NN), Nearest Subspace (NS) \cite{Lee2005-PAMI}, and Linear
Discriminant Analysis (LDA)
\cite{Belhumeur1997-PAMI}.\footnote{We do not list results on
PCA \cite{Turk1991-CVPR} as its performance is always below
that of Nearest Subspace.} Since these algorithms assume
pixel-accurate alignment, they are not expected to work well if
the test image is not well aligned with the training. In
Table~\ref{tab:MultiPIE-recognition}, we report the results of
these classical algorithms with three types of testing image
alignment: 1.\ alignment from the Viola and Jones' detector,
2.\ alignment via manually selected outer eye corners, and 3.\
the output of our iterative alignment algorithm. The
performance drop of the LDA algorithm on Multi-PIE reported
here seems to agree with that reported already in
\cite{Gross2008-FGR}.  All of the classical algorithms benefit
greatly from being paired with our iterative alignment
algorithm.

\begin{table}
\caption{Recognition rates on the Multi-PIE database for
different pairings of alignment and recognition stages.}
\vspace{-.2in} \centerline{
\begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c| }
\hline
\backslashbox{Rec.}{Align.}
& \multicolumn{3}{|c|}{Face Detector}
& \multicolumn{3}{|c|}{Manual}
& \multicolumn{3}{|c|}{Iterative Alignment}
\\
\hline
Session $\rightarrow$	& 2		&3			&4			& 2		&3			&4			& 2		&3			&4		\\
\hline
NS	& 30.8\%	& 29.4\%	& 24.6\%	& 77.6\%	& 74.3\%	& 73.4\%	& 84.5\%	& 82.3\%	& 81.4\% \\
\hline
NN	& 26.4\%	& 24.7\%	& 21.9\%	& 67.3\%	& 66.2\%	& 62.8\%	& 73.5\%	& 69.6\%	& 69.3\% \\
\hline
LDA	& 5.1\%		& 5.9\%		& 4.3\%		& 49.4\%	& 44.3\%	& 47.9\%	& 91.0\%	& 89.9\%	& 88.1\% \\
\hline
LBP	& 39.9\%	& 38.1\%	& 33.9\%	& 93.3\%	& 91.2\%	& 92.9\%	& {\bf 95.2\%}	& {\bf 94.7\%}	& {\bf 93.5\%} \\
\hline
SRC	& -- & -- & -- & -- & -- & -- & 93.9\%	& 93.8\%	& 92.3\% \\
\hline
\end{tabular}
\label{tab:MultiPIE-recognition} } \vspace{-.4in}
\end{table}

% LBP
We also compare our result to Local Binary Patterns (LBP)
\cite{Ahonen2006-PAMI}, a local appearance descriptor which is
able to capture fine details of facial appearance and texture.
Due to its robustness to variations in illumination, facial
expression, aging and other changes, LBP has achieved the
state-of-the-art face recognition performance in the scenario
when only one sample per person is used for training
\cite{Tan06facerecognition}. In this paper, we follow the same
steps as in \cite{Ahonen2006-PAMI} to construct an LBP
descriptor for each training and testing sample. The $80\times
60$ face region is first divided into a regular $10\times 10$
grid of cells, each of size $8\times 6$ pixels. Within each
cell, the histogram of 59 uniform binary patterns is then
computed, where the patterns are generated by thresholding 8
neighboring pixels in a circle of radius 2 using the central
pixel value. Finally, the local histograms are concatenated to
produce the global descriptor vector. As suggested in
\cite{Ahonen2006-PAMI}, the recognition is performed using a
nearest neighbor classifier with Chi square distance as the
distance measure and we report the recognition rates with the
same three types of input as before.  As shown in
Table~\ref{tab:MultiPIE-recognition}, although LBP achieves
competitive recognition rates given manually aligned training
and testing samples, demonstrating its robustness to moderate
misalignment, it still benefits from using the output of our
iterative alignment algorithm as the input. In addition, like
the other classical algorithms, the performance of LBP degrades
dramatically if it is applied directly to the output of a face
detector. This is notable given that LBP is often applied
without any special alignment in practice. Finally, we
attribute the improvement in performance of LBP over SRC in
this experiment to its robustness to illumination components
that cannot be linearly interpolated by the training set.

In addition, although our algorithm is not designed for
recognition when there is only a single gallery image per user,
we compare its performance with LBP within this setting for
completeness. For this experiment, we use the FERET dataset
\cite{phillips1998feret}, which contains five standard
partitions: `fa' is the gallery containing 1196 frontal images
of 1196 subjects, and `fb', `fc', `dup1' and `dup2' are four
sets of probe images. The testing sets differ from the training
in facial expression (`fb'), illumination (`fc'), aging (`dup1'
) and long aging (`dup2'). In fact, except for `fb', we notice
significant changes of illumination in all the other three test
sets. For the training, we again crop and normalize the face
region from each original image to an $80\times 60$ window
using manually marked eye coordinates \cite{Deng2010-PR}. In
Table~\ref{tab:FERET-recognition}, we report the performance of
our algorithm on the four test sets, with input directly
obtained from the Viola and Jones' detector, and the
performance of LBP with input obtained from both face detector
and manual alignment using the eye coordinates.

@@

\begin{figure}[t]
\centerline{
	\begin{tabular}{cc}
	\includegraphics[width=3in]{figures_pami/pami_roc_revision2.png} &
	\includegraphics[width=3in]{figures_pami/pami_roc2.png} \\
	{\small All algorithm with iterative alignment} & {\small Classical algorithms with manual alignment}
\\
	\end{tabular}
}
\caption{{\bf ROC curves} for subject validation on Multi-PIE database, (a) with iterative alignment,
and (b) for the classical algorithms with manual alignment (indicated by a subscript ``m'').}\label{fig:roc-multipie}
\vspace{-.4in}
\end{figure}

Figure \ref{fig:roc-multipie} plots the receiver operating
characteristic (ROC) curves, which are generated by sweeping
the threshold $t$ through the entire range of possible values
for each algorithm.\footnote{Rejecting invalid images not in
the entire database is much more difficult than deciding if two
face images are the same subject. Figure 8 should not be
confused with typical ROC curves for face similarity, e.g.,
\cite{PhillipsP2007}.} On the left we can see that the SCI
based recognition approach significantly outperforms the other
algorithms, including LBP, even when all algorithms are coupled
with our proposed iterative alignment.  In the right plot we
again see that classical algorithms, and even LBP, are very
sensitive to alignment.  Similar contrasts between our
algorithm and baseline algorithms were also observed for SRC in
\cite{Wright2009-PAMI}, though on much smaller datasets.

@@

\begin{figure}[t]
\centering
\includegraphics[scale=0.75,clip=true]{figures_pami/multipie_occ/occ10.png}
\includegraphics[scale=0.75,clip=true]{figures_pami/multipie_occ/occ20.png}
\includegraphics[scale=0.75,clip=true]{figures_pami/multipie_occ/occ30.png}
\includegraphics[scale=0.75,clip=true]{figures_pami/multipie_occ/occ40.png}
\includegraphics[scale=0.75,clip=true]{figures_pami/multipie_occ/occ50.png}
\vspace{-.25in}
\caption{{\bf Recognition under varying level of
random block occlusion.} The above row shows examples of occluded test images with occlusion level from 10\% to 50\%. Our method maintains high recognition rates up to 30\% occlusion:}
\vspace{3mm}
{\small
\begin{tabular}{|c|c|c|c|c|c| }
\hline
Percent occluded & 10\% & 20\% & 30\% & 40\% & 50\%  \\
\hline
\hline
Recognition rate & 99.6\% & 94.9\% & 79.6\% & 46.5\% & 19.8\% \\
\hline
\end{tabular}
}
\label{fig:multipie-occ-rec}
\vspace{-.4in}
\end{figure}
